---
title: "Class09"
author: "Matthew Chan (A18130675)"
format: pdf
---
## 1 Background
Objective: Your task is to explore their candy dataset to find out answers to these types of questions - but most of all your job is to have fun, learn by doing hands on data analysis, and hopefully make this type of analysis less frightening for the future! Let’s get started.

## 2 Importing candy data

```{r}
candyfile <- "candy-data.csv"
candy = read.csv(candyfile, row.names=1)
head(candy)
```

# 2.1 What is in the dataset

>Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` rows in this dataset

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

There are `r sum(candy$fruity)` fruity candy types in this dataset

```{r}
sum(candy$fruity)
```

# 2.2 What is your favorite candy?
```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="") |> 
  select(winpercent)
```

>Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

Haribo gold bears have a winpercent of 57.11974%

```{r}
candy |> 
  filter(row.names(candy)=="Haribo Gold Bears") |> 
  select(winpercent)
```

>Q4. What is the winpercent value for “Kit Kat”?

Kit Kat have a winpercent of 76.7686%

```{r}
candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

Tootsie Roll Snack Bars have a winpercent of 49.6535%

```{r}
candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The variable that looks to be on a different scale is winpercent.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero is for candy that is not chocolate and the one is for a candy considered a chocolate. 

## 3 Exploratory analysis
>Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
hist(candy$winpercent, breaks=20)
```
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_density(bins=12, fill= "lightblue", col="gray")
```


> Q9. Is the distribution of winpercent values symmetrical?

No, it doesnt look symmetrical it seems skewed to the right.

> Q10. Is the center of the distribution above or below 50%?

```{r}
ggplot(candy)+
  aes(winpercent) +
  geom_boxplot()

mean(candy$winpercent)
```

The center of the distribution is `r mean(candy$winpercent)` and is above 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Steps to solve this: 
1. find all chocolate candy in the dataset 
2. extract or find their winpercent values
3. calculate the mean of these values
4. find all fruit candy in the dataset and repeat

```{r}
choc.candy <- candy[candy$chocolate == 1,]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.candy <- candy[ candy$fruity == 1,]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

```{r}
mean(candy$winpercent[candy$chocolate == 1])

mean(candy$winpercent[candy$fruity == 1])
```

Chocolate candy is higher ranked than fruity candy with a winrate of `r mean(candy$winpercent[candy$chocolate == 1])` to `r mean(candy$winpercent[candy$fruity == 1])`

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```
Yes, because the p-value is less than 0.05

## 4 Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
inds <- order(candy$winpercent)
head(candy[inds, ], 5)
```

```{r}
candy |> arrange(winpercent) |> head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
inds <- order(candy$winpercent, decreasing = T)
head(candy[inds, ], 5)
```

```{r}
candy |> arrange(desc(winpercent)) |> head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=10, fig.width=7}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() + 
  ylab("") +
  theme_classic()
```

# 4.0.1 Time to add some useful color
```{r, fig.height=10, fig.width=7} 
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols) +
  ylab("") +
  theme_classic()
```

Now, for the first time, using this plot we can answer questions like:

> Q17. What is the worst ranked chocolate candy?

Nestle Smarties
> Q18. What is the best ranked fruity candy?

Starburst

## 5 Taking a look at pricepercent
We can use the **ggrepel** package for better label placement

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=10 )
```
The candy type highest ranked in terms of winpercent for the least money is Tootsie Roll Midgies with 45.73% winpercent for 0.011 price percent or Hershey's Kisses for 55.37545% winpercent to 0.093 price percent.
> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = T)
head( candy[ord,c(11,12)], n=5 )
```
The most expensive candy types in the dataset were Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate, with Nik L Nip being the least popular.

# 5.1 Optional
> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r, fig.height=10, fig.width=7}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```

```{r, fig.height=10, fig.width=7}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## 6 Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity, chocolate
> Q23. Similarly, what two variables are most positively correlated?

chocolate, winpercent

## 7 PCA Principal Component Analysis
```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```
The main results figure: the PCA score plot:

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols) + 
  labs(title="PCA Candy Space Map")
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
#library(plotly)
#ggplotly(p)
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The "loadings" plot for PC1. The original variables in the positive direction are fruity, pluribus, and hard. These make sense because they have a correlation. This relationship was previously highlighted in the correlation plot. 

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation),PC1)) +
  geom_col()
```

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The winning candies came from chocolate candy bars. They all show the correlations of how different candies are similar to each other, and then how those candies that are similar to each other have similar winpercents over other candies. 

















