---
title: "Class18"
author: "Matthew Chan (A18130675)"
format: pdf

---

## Background
Pertusis (a.k.a. whooping cough) is a common lung infection caused by bacteria B. Pertusis.

The CDC tracks the number of [reported data in the US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html): 
We can "scrape" this data with the **datapasta** package.

```{r}
cdc <- data.frame(data.table::data.table(
                                      year = c(1922L,1923L,
                                               1924L,1925L,1926L,1927L,1928L,
                                               1929L,1930L,1931L,1932L,1933L,
                                               1934L,1935L,1936L,1937L,
                                               1938L,1939L,1940L,1941L,1942L,
                                               1943L,1944L,1945L,1946L,1947L,
                                               1948L,1949L,1950L,1951L,
                                               1952L,1953L,1954L,1955L,1956L,
                                               1957L,1958L,1959L,1960L,1961L,
                                               1962L,1963L,1964L,1965L,1966L,
                                               1967L,1968L,1969L,1970L,
                                               1971L,1972L,1973L,1974L,1975L,
                                               1976L,1977L,1978L,1979L,1980L,
                                               1981L,1982L,1983L,1984L,
                                               1985L,1986L,1987L,1988L,1989L,
                                               1990L,1991L,1992L,1993L,1994L,
                                               1995L,1996L,1997L,1998L,1999L,
                                               2000L,2001L,2002L,2003L,
                                               2004L,2005L,2006L,2007L,2008L,
                                               2009L,2010L,2011L,2012L,2013L,
                                               2014L,2015L,2016L,2017L,
                                               2018L,2019L,2020L,2021L,2022L,
                                               2023L,2024L,2025L),
              cases = c(107473,164191,
                                               165418,152003,202210,181411,
                                               161799,197371,166914,172559,
                                               215343,179135,265269,180518,
                                               147237,214652,227319,103188,
                                               183866,222202,191383,191890,
                                               109873,133792,109860,156517,74715,
                                               69479,120718,68687,45030,
                                               37129,60886,62786,31732,28295,
                                               32148,40005,14809,11468,17749,
                                               17135,13005,6799,7717,9718,
                                               4810,3285,4249,3036,3287,
                                               1759,2402,1738,1010,2177,2063,
                                               1623,1730,1248,1895,2463,
                                               2276,3589,4195,2823,3450,4157,
                                               4570,2719,4083,6586,4617,
                                               5137,7796,6564,7405,7298,7867,
                                               7580,9771,11647,25827,25616,
                                               15632,10454,13278,16858,27550,
                                               18719,48277,28639,32971,
                                               20762,17972,18975,15609,18617,
                                               6124,2116,3044,7063,22538,21996)
            ))
```

> Q1 Plot of year over # of cases

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) +
  geom_line() + 
  geom_point() +
  labs(x = "Year", y = "Number of Cases", title = "Pertussis Cases by Year (1922-2025)")
```

> Q2 Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) + 
  aes(year, cases) +
  geom_point() +
  geom_line() + 
  
  geom_vline(xintercept = 1946, col = "#FD383e", lty=2) +
  geom_vline(xintercept = 1996, col = "#ed9ff3", lty=2) +
  geom_vline(xintercept = 2020, col = "#e39", lty=2) +
  labs(x = "Year", y = "Number of Cases", title = "Pertussis Cases by Year (1922-2025)")
```
Notice high number of cases before the wP vaccine and then a decline and stabilization, followed by an increase in 1996 after the introduction of the aP vaccine. Relatively stable from 1975 to 1996 and then unstable again.

> Q3 Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend? 

In the 1940s there was a high number of pertussis cases, but after the introduction of the wP vaccine in 1946 it dropped. After the switch to the aP vaccine in 1996, there was a small increase in the number of cases. Possible explanations for this increase in the 2000s is because of more sensitive PCR testing, vaccination hesistancy and distrust, and bacterial evolution of the B. Pertusis bacterium.

# Exploring CMI-PB data

[Computational Models of Immunity - Pertussis Boost](https://www.cmi-pb.org) project aims to provide the scientific community with this very information.

They make their data available via JSON format returning API. We can read this in R with the `read_json()` function from the **jsonlite** package:
```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

> Q8. Determine the age of all individuals at time of boost?

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

To analyze this data we need to first "join" (merge/link) the different tables so we have all the data in one place, not spread across different tables. 

We can use the `*_join()` family of functions from **dplyr** to do this
Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)

head(specimen)
head(ab_titer)
```


> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> What antigens are reported?

```{r}
table(abdata$antigen)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
The number of rows from the most recent is significantly higher than the previous two years, and is slightly less than if you were to combine the 2021 and 2022 datasets. however it is about half of the 2020 dataset.

Examine IgG Ab titer levels

```{r}
igg <- abdata |>
  filter(isotype == "IgG")
head(igg)
```
 
> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()  +
   xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q. Is there a difference between aP vs wP vaccinated individuals with these values?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

> Q. Is there a temprol response - i.e do values increase or decrease over time?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2) 
```

The values tend to increase initially and then decrease over time. 

# Focus on "PT" Pertussis Toxin antigen
```{r}
pt.igg.21 <- igg |> filter(antigen == "PT",
              dataset == "2021_dataset")

ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost,
      MFI_normalised,
      col=infancy_vac,
      group = subject_id)+
  geom_point(alpha = 0.4) +
  geom_line(alpha = 0.4) +
  geom_smooth(aes(group = infancy_vac), se = T, span = 0.5)
  geom_vline(xintercept = 14, lty=2)

```
> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens that show differences in the level of IgG antibody titers recognizing them over time are PT and FIM 2/3 because they get wider with time and have boxes that have higher MFI in the middle visits.. However, PRN and FHA also show increases in the middle, with TT, OVA, and DT having the weakest changes. These antigens may elicit stronger responses.







