---
title: "Class07"
author: "Matthew Chan (A18130675)"
format: pdf
---

## Background 

Today we wuill begin our exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**

Let's make up some input data for clustering where we know what the natural "clusters" are.

The function `rnorm()` can be useful here. It returns a bell curve out of a sample n, with mean and sd as things you can change.

```{r}
hist( rnorm(5000, mean=10) )
```

> Q. Generate 30 random numbers centered at 3 and another 30 centered at -3

```{r}

tmp <- c( ( rnorm(30, mean=3) ), rnorm(30, mean=-3) ) #combines both rnorm distributions

x <- cbind(x=tmp, y=rev(tmp)) #compiles to column

plot(x)
```

\## K-means clustering

The main function in "base R" for K-means clustering is called surprisingly `kmeans()`

```{r}
km <- kmeans(x, centers= 2)
km
```

> Q. What component of the results object details the cluster sizes

```{r}
km$size
```

> Q. What component of the results object details the cluster centers?

```{r}
km$centers
```

> What component of the results object details the cluster membership vector( ie. our main result of which points lie in which cluster) ?

```{r}
km$cluster
```

> Q. Plot our clustering results with points colored by cluster and also add the cluster centers as new points colored blue?

```{r}
plot( x, col=km$cluster) #plots the color, uses the membership to color
points(km$centers, col="blue", pch=15) #adds points to an existing plot
```

> Q. Run `kmeans()` again and this time produce 4 clusters (and call your result object `k4`) and make a results figure like above?

```{r}

k4 <- kmeans(x, centers = 4)
plot(x, col=k4$cluster)
points(k4$centers, col="red", pch=15)
```

The metric

```{r}
km$tot.withinss
k4$tot.withinss
```

> Q. Let's try different number of K (centers) from 1 to 30 and see what the best result is

```{r}
ans <- NULL 
for(i in 1:30) {
  ans <- c(ans, kmeans(x, centers =i)$tot.withinss)
}

ans
```

```{r}
plot(ans, typ="o") #gives a scree plot. it gives you the most with the least number of centers. you want the elbow part of the graph. The problem with k is that it is self fulfilling
```

**Key-Point:** K-means will impose a clustering structure on your data even if it is not there - it will always give you the answer you asked for even if that answer is silly.

## Hierarchieal Clustering

The main function for Hierarchial Clustering is called `hclust()`. Unlike `kmeans()` (which does all the work for you) you cant just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

To extract our cluster membership vector from a `hclust()` result object we have to "cut" our tree at a given height to yield separate "groups/branches"

```{r}
plot(hc)
abline(h=8, col="red", lty=2)
```

To do this we use the `cutree` function on our `hclust()` object:

```{r}
grps <- cutree(hc, h=8)
grps
```

```{r}
table(grps, km$cluster)
```

## PCA of UK food data

Import the dataset of food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

One solution to set the row names is to do it by hand

```{r}
rownames(x) <- x[,1]
x
```

To remove the first column I can use the minus index trick.
```{r}
x <- x[,-1]
```

A better way to do this is to set the row names to the first column  with `read.csv()`

```{r}
x <- read.csv(url, row.names = 1)
```

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

## Spotting major differences and trends
Is difficult even in this wee 17D dataset

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

> Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

if a given plot is off of the diagonal then there will be a difference. for ireland there are many points that are off the diagonal than the other countries

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}
library(pheatmap)
pheatmap(x)
```

## PCA to the rescue 

The main PCA function in "base R" is called `prcomp()`  This function wants the transpose of our food data as input (i.e. the foods as columsna dn the countries as rows). 

```{r}
pca <- prcomp(t(x))
```

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

To make one of our main PCA result figures we turn to `pca$x` the scores along our new PCs . This is called "PC plot" or "score plot" or "Ordination plot"

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
```


```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1, PC2) +
  geom_point(col=my_cols)
  
```

The second major result figure is called a "loadings plot" or "variable contributions plot" or "weight plot"

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```


